### 2025-02-10 paper/sections/other/constraints.qmd
progress

@@

# Linear Model

First, we define the mutability constraints and load the data and model:

@@

@@

```{julia}
using CounterfactualExplanations.Convergence
obj = EnergyDifferentialObjective(lambda=[1.0,1.0,0.0])
# obj = VanillaObjective()
generator = GenericGenerator()
opt_state = Flux.setup(Descent(), model)
conv = MaxIterConvergence()
```

@@

Then we train the model:

>>>>

---
execute: 
  eval: false
---

@@

## Details on Encoding Mutability Constraints {#sec-app-constraints}

In @sec-constraints we explained that to avoid penalizing implausibility that arises due to mutability constraints for any feature $d$, we impose a point mass prior on $p(\mathbf{x})$ for that features. This is best explained through a simple example involving a binary linear classifier.

:::{exm-constraints}

Let $\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}$, with $y \in \{1,2\}$. Then for $y^+ \sim p(y)=2$ we have:

$$
\nabla_\theta \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) = \begin{pmatrix} 0 & 0 \\ \mathbf{x}^\prime[1] - \mathbf{x}[1] & \mathbf{x}^\prime[2] - \mathbf{x}[2] \end{pmatrix}
$$ {#eq-grad}

Contrastive divergence pushes down (up) on the coefficient $\Theta[y^+,d]$ by the amount that feature $d$ that is lower (higher) for counterfactuals than samples. This has the intended effect of aligning counterfactual explanations with the data, as discussed above.

Suppose, for example, that feature $d=1$ represents *age* and our model predicts the likelihood of credit default: $y \in \{1:=\text{'no default'},2:=\text{'default'}\}$. Then if $\mathbf{x}^\prime[1] < \mathbf{x}[1]$, contrastive divergence will push down on the corresponding coefficient, $\Theta[\text{'default'},\text{'age'}]$. Intuitively, this has the effect of decreasing the effect of *age* that the model attributes to the log odds of default, i.e. the model learns to higher default risk younger individuals. But if *age* is immutable, then this corresponds to penalizing younger individuals for implausibility that arises due to a feature that they can change about themselves. This is precisely what we want to avoid, and therefore we impose no penalty at all.

:::

@@

## Gradient

```{julia}
X1 = ce_data.X[:,ce_data.output_encoder.labels.==1]
X2 = ce_data.X[:,ce_data.output_encoder.labels.==2]
factual = X1[:,1]
target = 2
target_enc = Flux.onehotbatch([target], [1,2])
neighbor = X2[:,1]
```

## Full Training Loop

@@

```{julia}
using CounterfactualExplanations.Convergence
obj = EnergyDifferentialObjective(lambda=[1.0,1.0,0.0])
generator = GenericGenerator()
opt_state = Flux.setup(Descent(), model)
conv = MaxIterConvergence()
```

@@
